---
title: "Predictive Text Model — Milestone Report"
subtitle: "Exploratory analysis and modeling plan for a next-word prediction Shiny app"
format:
  html:
    toc: true
    embed-resources: true
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(data.table)
library(ggplot2)
library(scales)

out_dir <- "../data/processed"
raw_stats          <- as.data.table(readRDS(file.path(out_dir, "raw_stats.rds")))
words_per_line      <- readRDS(file.path(out_dir, "eda_words_per_line.rds"))
unigram_freq        <- readRDS(file.path(out_dir, "eda_unigram_freq.rds"))
coverage_summary     <- readRDS(file.path(out_dir, "eda_coverage_summary.rds"))
coverage_curve       <- readRDS(file.path(out_dir, "eda_coverage_curve.rds"))
bigram_freq          <- readRDS(file.path(out_dir, "eda_bigram_freq.rds"))
trigram_freq         <- readRDS(file.path(out_dir, "eda_trigram_freq.rds"))
source_summary       <- readRDS(file.path(out_dir, "eda_source_summary.rds"))
model_sizes          <- readRDS(file.path(out_dir, "model_sizes.rds"))
grid_results         <- readRDS(file.path(out_dir, "model_grid_results.rds"))
lambda_results       <- readRDS(file.path(out_dir, "model_lambda_results.rds"))
comparison           <- readRDS(file.path(out_dir, "model_comparison.rds"))
model_config         <- readRDS(file.path(out_dir, "model_config.rds"))
test_eval            <- readRDS(file.path(out_dir, "model_eval_test.rds"))
```

## What this project is about

The goal is a Shiny app where a user types a phrase and the app suggests the
most likely next word — similar to the predictive keyboard on a smartphone.
This report covers the first stage: understanding the raw text data, cleaning
it, and building (and sanity-checking) a first, basic prediction model. It's
written for a general audience — no statistics background required.

## The data

The underlying data are three large collections of English text, scraped from
the web: Twitter posts, blog posts, and news articles.

```{r raw-stats-table}
raw_stats[, .(Source = source, `Size (MB)` = file_size_mb,
              `Lines` = comma(n_lines), `Words` = comma(n_words),
              `Avg. words/line` = avg_words_per_line)]
```

Blogs and news articles run around 34–42 words per entry on average, while
tweets are much shorter (about 13 words), reflecting the platform's length
limit. Together the three files total over 100 million words — far more than
needed to build a reliable model, so we work with a large random sample
(1,000,000 lines, allocated proportionally across the three sources) rather
than the full data. This keeps computation fast while still being
representative. That sample is further split into three pieces: 80% for
building the model's word-count tables, 10% held out to tune settings, and a
final 10% touched only once, at the end, to report honest performance
numbers — the same discipline used for any predictive model, so we don't fool
ourselves about how well it generalizes.

```{r words-per-line-plot}
#| fig-cap: "Distribution of words per line by source (sample, capped at 80 words for readability)"
ggplot(words_per_line[N <= 80], aes(x = N)) +
  geom_histogram(binwidth = 2) +
  facet_wrap(~source, scales = "free_y") +
  labs(x = "Words per line", y = "Number of lines") +
  theme_minimal()
```

## Cleaning the text

Raw web text is messy. Before counting words, each line is: lower-cased;
stripped of URLs, @-handles, and hashtag symbols; stripped of punctuation and
emoji; and split into individual word "tokens." Pure numbers are dropped,
since they aren't useful things to *predict* as a next word.

The text also contains profanity, which we don't want the app to ever
suggest. We filter out matches against a list of roughly 500 common English
profane terms before any word counts are used for prediction. In this sample,
profanity accounted for about 0.2% of all words — a small but real share, and
worth screening out.

## What the words look like

A small number of words dominate everyday English. In this sample, just
**`r coverage_summary[coverage == 0.5, n_words_needed]` distinct words**
account for half of all the words used, and the top 10 alone are shown below.

```{r top-words-plot}
#| fig-cap: "15 most frequent words in the sample"
top_words <- head(unigram_freq[order(-N)], 15)
ggplot(top_words, aes(x = N, y = reorder(word, N))) +
  geom_col() +
  labs(x = "Occurrences", y = NULL) +
  scale_x_continuous(labels = comma) +
  theme_minimal()
```

This "a few words do most of the work" pattern (Zipf's law) is useful for
building an efficient model: we don't need to keep every rare word to cover
most of what people actually type.

```{r coverage-table}
coverage_summary[, .(`Share of all words covered` = percent(coverage),
                      `Distinct words needed` = comma(n_words_needed))]
```

```{r coverage-plot}
#| fig-cap: "Cumulative share of all word occurrences covered by the N most common distinct words"
ggplot(coverage_curve, aes(x = rank, y = cum_share)) +
  geom_line() +
  scale_x_log10(labels = comma) +
  scale_y_continuous(labels = percent) +
  labs(x = "Number of distinct words (most frequent first, log scale)",
       y = "Cumulative share of all word occurrences") +
  theme_minimal()
```

Word pairs and triples ("n-grams") show similarly strong patterns — mostly
functional phrases like "of the" or "one of the" rather than distinctive
content:

::: {layout-ncol=2}
```{r top-bigrams}
head(bigram_freq[order(-N)], 10)[, .(`Word pair` = ngram, Count = comma(N))]
```

```{r top-trigrams}
head(trigram_freq[order(-N)], 10)[, .(`Word triple` = ngram, Count = comma(N))]
```
:::

## The prediction model

The prediction approach is a classic **n-gram model**: count how often each
word follows the previous 1, 2, or 3 words in the training text, and use
those counts to rank candidate next words. Concretely, we built four lookup
tables — for single words, pairs, triples, and quadruples of words — from the
cleaned, profanity-filtered sample.

**Handling phrases we've never seen before.** No matter how much text we
train on, users will type combinations of words that never appeared in the
training data. We compared two standard strategies for this:

- **Stupid Backoff**: try to match the user's last three words; if that exact
  three-word phrase was never observed, fall back to the last two words, then
  the last word, then (if all else fails) simply suggest the most common
  words overall.
- **Linear interpolation**: rather than committing to a single "best"
  context length, blend evidence from all the context lengths that *do*
  match (single word, pair, triple, and quadruple), weighting longer,
  more specific contexts more heavily.

Both are simpler and far more memory-efficient than more statistically
rigorous smoothing methods (such as Katz backoff with Good-Turing
discounting, or Kneser-Ney smoothing), at the cost of not producing
mathematically "proper" probabilities — an acceptable tradeoff for a
responsive, lightweight app, though it does mean the "perplexity" figure
below should be read as a rough diagnostic rather than a textbook one.

To keep the model small enough to run comfortably in a memory-constrained
Shiny deployment, we also tune two pruning settings — the minimum number of
times a phrase must appear to be kept, and how many candidate next-words to
keep per phrase — on the tuning slice:

```{r grid-search-table}
grid_results[order(-dev_accuracy), .(`Min. count` = min_count, `Candidates kept per phrase` = top_k,
             `Tuning accuracy` = percent(dev_accuracy, accuracy = 0.1), `Table rows` = comma(n_rows))]
```

Requiring a phrase to appear at least twice turns out to matter more than how
many candidates we keep per phrase — beyond keeping the top 5, extra
candidates rarely help, because most specific 2-3 word phrases simply don't
have many different continuations that occur more than once in this amount
of data. We locked in **minimum count `r model_config$min_count`, top
`r model_config$top_k` candidates per phrase** based on this.

With pruning settings fixed, we compared Stupid Backoff against linear
interpolation (with interpolation weights also chosen on the tuning slice):

```{r comparison-table}
comparison[, .(Model = model, `Tuning accuracy` = percent(dev_accuracy, accuracy = 0.1))]
```

Interpolation won, so that's what the model uses going forward — blending
evidence across context lengths captures useful signal that strict backoff
throws away whenever it commits to a single context length.

```{r model-size-table}
model_sizes[, .(`N-gram table` = table, `Rows kept` = comma(n_rows),
                 `Size in memory (MB)` = round(size_mb, 1))]
```

The resulting four tables together use about
**`r round(sum(model_sizes$size_mb))` MB** of memory — comfortably within the
4GB available on the deployment server.

**Is it any good?** We report final numbers on the test slice — data
touched only for this one measurement, never used to choose any setting
above, to keep the numbers honest. Taking the first three words of each test
sentence and asking the model to guess the fourth, the correct word appeared
in the model's top 3 guesses **`r percent(test_eval$top3_accuracy, accuracy = 0.1)`**
of the time (pseudo-perplexity: **`r round(test_eval$pseudo_perplexity)`**,
lower is better; again, a rough diagnostic rather than a directly comparable
textbook figure, since Stupid Backoff/interpolation scores aren't a properly
normalized probability distribution). That's a modest improvement over our
first pass, driven mostly by fixing a bug in how the model chose between
context lengths and by blending contexts instead of picking just one — a
reasonable baseline for a model with no linguistic knowledge built in, and
still room to improve, principally by training on more data than the current
1,000,000-line sample (see next steps).

## Plan for the Shiny app and next steps

The app itself will be simple from the user's point of view: a text box where
someone types a partial sentence, and up to three suggested next words appear
as they type — similar to a phone's predictive keyboard.

Behind the scenes, next steps are to:

- **Train on more data** — the current model uses 1,000,000 sampled lines;
  the biggest known weakness (many 3-word phrases in new text were never
  seen during training) would likely improve most from simply learning from
  more of the >100 million words available, now that the pipeline runs
  comfortably in memory. Not done in this pass, to keep runtime and model
  size predictable while other improvements were tested.
- **Consider more principled smoothing** — Katz backoff with Good-Turing
  discounting or Kneser-Ney smoothing are more statistically rigorous than
  Stupid Backoff/interpolation, at real added implementation complexity;
  worth revisiting if accuracy gains from more data plateau.
- **Consider a neural language model** as a further-out option — could
  capture more context and meaning than word counts, but adds heavy new
  dependencies, longer training time, and latency considerations for a
  live-typing app on a memory-constrained server, so it's not planned unless
  simpler improvements stop paying off.
- **Speed up lookups** — pre-index the tables so a prediction returns
  effectively instantly as the user types.
- **Build the app** — wire the model into a Shiny text input with live
  suggestions, and deploy it within the memory budget confirmed above.
- **Basic quality guardrails** — keep the profanity filter in place so the
  app never suggests offensive words, and monitor a few example phrases by
  hand before wider testing.
